% Simple article, but with some useful ApJ things included
% Created by AYQH, 18 June 2016
\documentclass[12pt, letterpaper, preprint]{aastex}
\bibliographystyle{apj}
\usepackage{float, bm, graphicx, subfigure, amsmath, morefloats}
\usepackage{color}
\usepackage{tabularx}

% useful macros
\newcommand{\angstrom}{\mbox{\AA}}

% for The Cannon
\newcommand{\tc}{\textsl{The~Cannon}} 
\newcommand{\apogee}{APOGEE}
\newcommand{\aspcap}{ASPCAP}
\newcommand{\lamost}{LAMOST}
\newcommand{\apokasc}{APOKASC}
\newcommand{\kepler}{\textsl{Kepler}}
\newcommand{\ulyss}{\textsl{ULySS}}
\newcommand{\teff}{\mbox{$\rm T_{eff}$}}
\newcommand{\feh}{\mbox{$\rm [Fe/H]$}}
\newcommand{\mh}{\mbox{$\rm [Fe/H]$}}
\newcommand{\alphafe}{\mbox{$\rm [\alpha/M]$}}
\newcommand{\logg}{\mbox{$\rm \log g$}}
\newcommand{\cm}{\mbox{$\rm [C/M]$}}
\newcommand{\nm}{\mbox{$\rm [N/M]$}}
\newcommand{\starlabel}{\ell}
\newcommand{\starlabelvec}{\set{\starlabel}}

\begin{document}

\title{Written Report}
\author{Anna Ho}

\begin{abstract}

  I have worked on two methodologies. Methodology \#1: data-driven
  spectral modeling (The Cannon). Code here. Implementation:
  label transfer from APOGEE to LAMOST (Ho et al. 2016a, submitted),
  masses \& ages from LAMOST spectra (Ho et al. 2016b, submitted),
  individual element abundances from LAMOST spectra (Ho et al. 2016c,
  submitted). 
  Methodology \#2: On-The-Fly (OTF) mapping with the VLA (developed
  by Kunal Mooley for his PhD thesis). Code here. Implementation:
  follow up imaging in the LIGO field (worked on this over the summer).
  Implementation: follow up unidentified x-ray transients discovered
  by MAXI (worked on the proposal for this over the summer, we submitted this,
  will be my project with Shri starting in the fall). 

\end{abstract}

\section{The Cannon}

\subsection{New labels from LAMOST spectra}

\subsection{Mass \& age from LAMOST spectra}

\subsection{Individual element abundances from LAMOST spectra}

\section{On-The-Fly Mapping with the VLA}

\subsection{LIGO Follow-Up in the Gravitational Wave Era}

\subsection{Following up unidentified MAXI transients}

\end{document}


